\begin{thebibliography}{11}

\bibitem{uomlogo}
University of Manchester logo from Wikipedia by source, fair use\newline
\url{https://en.wikipedia.org/w/index.php?curid=43485475}\newline
Uploaded \nth{6} August 2014, Accessed \nth{11} April 2016

\bibitem{iprover}
Korovin, Konstantin. ``iProver-an instantiation-based theorem prover for first-order logic (system description).'' \textit{Automated Reasoning.} Springer Berlin Heidelberg, 2008. 292-298.

\bibitem{dpll}
Davis, Martin, George Logemann, and Donald Loveland. ``A machine program for theorem-proving.'' \textit{Communications of the ACM} 5.7 (1962): 394-397.

\bibitem{depqbf}
Lonsing, Florian, and Armin Biere. ``DepQBF: A dependency-aware QBF solver.'' \textit{Journal on Satisfiability, Boolean Modeling and Computation} 7 (2010): 71-76.

\bibitem{qbf2epr}
Seidl, Martina, Florian Lonsing and Armin Biere. ``qbf2epr: A Tool for Generating EPR Formulas from QBF.'' \textit{PAAR@ IJCAR.} 2012.
  
\bibitem{cookstheorem}
Cook, Stephen A. ``The complexity of theorem-proving procedures.'' \textit{Proceedings of the third annual ACM symposium on Theory of Computing.} ACM, 1971.

\bibitem{savitch}
Savitch, Walter J. ``Relationships between nondeterministic and deterministic tape complexities.'' \textit{Journal of computer and system sciences} 4.2 (1970): 177-192.

\bibitem{harrylewis}
Lewis, Harry R. ``Complexity results for classes of quantificational formulas.'' \textit{Journal of Computer and System Sciences} 21.3 (1980): 317-353.

\bibitem{skolemization}
Skolem, Thoralf. ``Logisch-Kombinatorische Untersuchungen {\"u}ber die Erf'{\"u}llbarkeit oder Beweisbarkeit Mathematischen S{\"a}tze nebst einem Theoreme {\"u}ber Dichte Mengen.'' Translation sourced from: Van Heijenoort, Jean. \textit{From Frege to G{\"o}del: a source book in mathematical logic, 1879-1931.} Vol. 9. Harvard University Press, 1967.

\bibitem{backdoorsets}
Samer, Marko, and Stefan Szeider. ``Backdoor sets of Quantified Boolean Formulas.'' \textit{Journal of Automated Reasoning} 42.1 (2009): 77-97.

\bibitem{ocaml}
OCaml homepage\newline
\url{https://ocaml.org}\newline
Accessed \nth{18} April 2016.

\bibitem{qdimacs}
QDIMACS grammar definition\newline
\url{http://www.qbflib.org/qdimacs.html}\newline
Released \nth{21} December 2005, accessed \nth{18} April 2016.

\bibitem{qbflib}
QBFLIB homepage\newline
\url{http://qbflib.org}\newline
Accessed \nth{18} April 2016.

\bibitem{tptp}
TPTP homepage\newline
\url{http://www.cs.miami.edu/\~tptp/}\newline
Accessed \nth{18} April 2016.

\end{thebibliography}
